European or American? A re-examination of the transatlantic influence over the IMF

### 2025-07-31 manuscript/manuscript.qmd
Add rearranged version

@@

European influence reflects a strategic combination of formal power and coordinated action. EU countries collectively control over 30% of IMF voting rights, and if they were to unify, they could potentially counterbalance US veto power [@brandner2009a; @leech2005a; @smaghi2004]. This is reinforced by the long-standing transatlantic norm allocating IMF leadership to a European national [@kahler2001]. Through platforms such as EURIMF and SCIMF, EU members have harmonised policy positions and influenced loan design. During the Eurozone crisis, France and Germany pushed for IMF involvement to stabilise the region while embedding austerity-oriented conditionality that protected their banking systems [@hodson2014; @kickert2019a; @veron2016]. Lending terms were shaped to reflect creditor priorities, with evidence that conditionality was both intensified or relaxed based on geopolitical alignment [@cain2022; @dreher2015; @lutz2014]. Beyond Europe, the UK and France have used IMF concessional lending in Africa to maintain post-colonial ties and commercial access [@moyo2024; @stone2004].

@@

While several studies examine European involvement in IMF operations, most focus on individual cases or regional crises, lacking generalisable conclusions. Existing literature often treats Europe as a secondary actor or subsumes it under undifferentiated “Western” influence, thereby overlooking intra-Western variation. By analytically separating US and European roles, our study addresses these gaps and offers a comparative framework for understanding divergent patterns of influence within global financial governance.

>>>>

Compared to the extensive literature on US dominance in the IMF, research on European influence remains more limited and fragmented. Nevertheless, a growing body of scholarship highlights that European countries do possess considerable institutional and structural advantages that enable them to shape lending outcomes. First, EU countries collectively hold over 30% of IMF voting shares—greater than the US—and, if coordinated, could theoretically counterbalance American veto power [@brandner2009a; @leech2005a; @smaghi2004]. Second, Europe has enjoyed long-standing leadership within the Fund: all IMF Managing Directors since the institution's founding have been European nationals [@kahler2001]. Third, coordination mechanisms such as EURIMF and SCIMF facilitate policy alignment among EU members and amplify their collective voice. During the Eurozone crisis, for instance, France and Germany lobbied for IMF involvement, embedding austerity-oriented conditionality that aligned with creditor interests and protected domestic banking sectors [@hodson2014; @kickert2019a; @veron2016]. Scholars have also documented European influence beyond the continent—for example, in the use of IMF concessional lending to sustain post-colonial relationships in Africa, particularly by the UK and France [@moyo2024; @stone2004]. These studies suggest that European actors, when effectively coordinated, are capable of steering lending priorities.

At the same time, existing literature raises questions about the consistency and effectiveness of European influence. Internal differences—such as the often-cited contrast between France’s growth-oriented approach and Germany’s fiscal conservatism—along with the dispersion of EU members across multiple IMF constituencies, are said to limit coherence and reduce policy impact [@smaghi2004; @miescu2014; @dreher2015]. While some proposals have called for the consolidation of EU seats to strengthen European influence, critics argue this may entrench the dominance of larger EU states without necessarily enhancing representativeness [@leech2005a; @copelovitch2010a]. Moreover, concerns about Europe’s voting power exceeding its global economic and demographic weight have generated debates about fairness and legitimacy, particularly from developing countries [@woodward2024]. These institutional tensions may place constraints on Europe’s ability to act decisively, though the extent of such limitations remains contested.

Although many studies document specific cases of European involvement—especially during regional crises—they tend to be case-based, focus on intra-European dynamics, or treat Europe as a secondary actor. In addition, several key works either overlook European influence altogether or conflate it with broader “Western” alignment [@lipscy2018; @barro2005]. As a result, generalisable, cross-national assessments of Europe’s role in shaping IMF outcomes remain scarce. This paper addresses that gap by analytically disaggregating US and European influence within a unified empirical framework. In doing so, it contributes to a more nuanced understanding of transatlantic dynamics in global financial governance—highlighting that Western influence may not be as unified or symmetrical as often assumed.

@@

European influence reflects a strategic combination of formal power and coordinated action. Lending terms were shaped to reflect creditor priorities, with evidence that conditionality was both intensified or relaxed based on geopolitical alignment [@cain2022; @dreher2015; @lutz2014].

@@
